---
title: "PA1_Template"
author: "Jigar Patel"
date: "December 8, 2018"
output:
  md_document
 
---


# Reproducible Research

Loading packages :

```{r echo = TRUE}
  library("dplyr")
  library("lubridate")
  library("data.table")
  library("tibble")
  library("ggplot2")
  library("lattice")
```

1. reading the data file using fread and tbl_df, also cleaning up the date column simultaneously using mutate

```{r echo = TRUE}
    rawdata<- tbl_df(fread("activity.csv", na.strings = "NA") ) %>%  mutate(date = ymd(date))
```



2.  filtering out NA values

```{r echo = TRUE}
    data <- filter(rawdata, steps != 'NA')
```



3. aggregating data to get total steps each day

```{r echo = TRUE}
  aggregated_totalsteps <- aggregate(x =data[c("steps","interval")], by = list(data$date),FUN = sum)
```



4. creating histogram of total steps per day using qplot

```{r echo = TRUE}
qplot(aggregated_totalsteps$steps, geom="histogram",binwidth= 5000, main = "Histogram for Steps", xlab = "Total Steps Per Day")
```



5.  mean steps

```{r echo = TRUE}
Meansteps <- mean(aggregated_totalsteps$steps)
```



6.  median steps

```{r echo = TRUE}
Mediansteps <- median(aggregated_totalsteps$steps)
```

## What is the average daily activity pattern?

7. aggregrating data by interval for time series plot

```{r echo = TRUE}
Byinterval <- aggregate(data$steps, by = list(data$interval), mean)
colnames(Byinterval)<-c("intervalID", "Mean steps")
```



8.  Time Series plot

```{r echo = TRUE}
plot(Byinterval$intervalID,Byinterval$`Mean steps`,type="l",xlab="5-Minute Interval",ylab="Mean Steps Taken",
     
     main="Steps during the Day")
```

9. Interval with most activity

```{r echo = TRUE}
Byinterval$intervalID[which.max(Byinterval$`Mean steps`)]
```


## Imputing missing values


10. Let us re-read the raw data

```{r echo = TRUE}
data2<- tbl_df(fread("activity.csv", na.strings = "NA") ) %>%  mutate(date = ymd(date))
```

11. How many missing values

```{r echo = TRUE}
sum(!complete.cases(data2))
```

12. Replacing missing values with mean from that
 specific interval (Ex: Missing value in interval
 5 replaced by mean step value in interval 5

```{r echo = TRUE}
summary <- data2[complete.cases(data2), ] %>% group_by(interval) %>% summarise(mean=mean(steps));
incomplete <- data2[!complete.cases(data2),];
updated <- merge(x=incomplete, y=summary, by="interval", all.x=T);
updated$steps <- updated$mean;
imputed <- rbind(data2[complete.cases(data2), ], select(updated, steps, date, interval))
```

13. no more NA values

```{r echo = TRUE}
sum(!complete.cases(imputed))
```

14. aggregating data2 to get total steps each day

```{r echo = TRUE}
aggregated_totalsteps2 <- aggregate(x =imputed[c("steps","interval")], by = list(imputed$date),FUN = sum)
```


15. creating the new histogram of total steps per day using qplot

```{r echo = TRUE}
qplot(aggregated_totalsteps2$steps, geom="histogram",binwidth= 5000, main = "Histogram for Steps", xlab = "Total Steps Per Day")
```



16. new mean steps

```{r echo = TRUE}
Meansteps2 <- mean(aggregated_totalsteps2$steps)
```



17. median steps

```{r echo = TRUE}
Mediansteps2 <- median(aggregated_totalsteps2$steps)
```


## Are there differences in activity patterns between weekdays and weekends?

18. Creating a new column that classifies date column into weekend/weekday

```{r echo = TRUE}
imputed$day<-ifelse(as.POSIXlt(as.Date(imputed$date))$wday%%6==0,
                         "weekend","weekday")

imputed$day<-factor(imputed$day,levels=c("weekday","weekend"))
```

19. Creating a subset data contains steps per interval, then creating a two panel chart

```{r echo = TRUE}
WeekdayWeekEndData <- aggregate(steps~interval+day,imputed,mean)
xyplot(steps~interval|factor(day), data=WeekdayWeekEndData,aspect=1/2, type="l")
```

The plots show that there is difference in movement activity between weekdays and weekends